% =============================================================================
%  Theory of Generated Space (TGP): Core
%  A self-contained preprint defining axioms, substrate, effective field,
%  principal equations, and the status of theorems and open problems.
%
%  Mateusz Serafin,  April 2026
%  To be published on Zenodo.
% =============================================================================
\documentclass[11pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{mathtools}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{longtable,booktabs,tabularx,array}
\usepackage[shortlabels]{enumitem}
\usepackage{xcolor}
\usepackage[protrusion=true,expansion=false]{microtype}
\geometry{margin=2.5cm}

% --- TGP symbols -------------------------------------------------------------
\renewcommand{\Phi}{\varPhi}
\newcommand{\PhiZero}{\varPhi_{0}}
\newcommand{\Nzero}{N_{0}}
\newcommand{\Ffield}{\mathcal{F}}
\newcommand{\lP}{\ell_{\mathrm{P}}}
\newcommand{\Zp}{\mathbb{Z}_{2}}
\newcommand{\DD}{\mathcal{D}}
\newcommand{\NN}{\mathcal{N}}

% --- Theorem environments ----------------------------------------------------
\newtheorem{axiom}{Axiom}
\newtheorem{definition}[axiom]{Definition}
\newtheorem{theorem}[axiom]{Theorem}
\newtheorem{proposition}[axiom]{Proposition}
\newtheorem{lemma}[axiom]{Lemma}
\newtheorem{corollary}[axiom]{Corollary}
\newtheorem{remark}[axiom]{Remark}
\newtheorem{hypothesis}[axiom]{Hypothesis}

\title{%
  \textbf{Theory of Generated Space (TGP):}\\[4pt]
  \large A minimal core of axioms, substrate, and effective field%
}
\author{Mateusz Serafin\\[2pt]
\small\textit{Independent researcher}\\[-2pt]
\small\texttt{environmenstefan@gmail.com}}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
We present the core of the \emph{Theory of Generated Space} (TGP):
a scalar-field theory in which space is not a background arena but is
\emph{constituted} by mass-energy sources through a shared field $\Phi$.
The theory is built on four axioms
(absolute reference state $\Nzero$, $\Zp$-symmetric relational substrate,
matter as source, instability of $\Nzero$) and a single dynamical degree of freedom.
From these inputs we derive:
(i) a unique second-order kinetic operator with coefficient $\alpha=2$,
supplied directly by the Ginzburg--Landau substrate bond
in~\eqref{eq:H-Gamma} (continuum form in App.~B,
Prop.~\ref{prop:substrate-action}) and confirmed by a classification
theorem within conditions (C1)--(C3) of Sec.~\ref{sec:field};
(ii) an effective metric whose temporal factor is the normalised
substrate-potential density $g_{tt}=-c_{0}^{2}\cdot 12\,U(\Phi)/(\gamma\PhiZero^{2}\psi^{4})
=-c_{0}^{2}(4-3\psi)/\psi$ (with $\psi=\Phi/\PhiZero$), reproducing
$\gamma_{\mathrm{PPN}}=\beta_{\mathrm{PPN}}=1$ exactly at 1PN with explicit,
falsifiable 2PN deviations ($|\Delta g_{tt}|_{2\mathrm{PN}}=\tfrac{5}{6}U_{\mathrm{N}}^{3}$);
(iii) three interaction regimes (long-range attraction, intermediate repulsion,
short-range confinement) from a single non-linear potential $U(\Phi)$;
(iv) identically luminal gravitational-wave propagation
$c_{\mathrm{GW}}=c_{0}$, consistent with GW170817.
TGP is \emph{not} a scalar-tensor theory: the metric is algebraically determined
by $\Phi$, and the action contains no Einstein--Hilbert term.
This document defines the minimal, self-contained core; applications
(superconductivity, lepton masses, black-hole shadows, dark energy) are
developed in companion papers that cite this reference.
We state twenty-one proven or numerically verified results and ten
explicit open problems, so that the theory is falsifiable and extendable
on a well-defined basis.
\end{abstract}

\noindent\textbf{Keywords:}
generated space, emergent gravity, scalar field, $\Zp$ substrate,
PPN parameters, effective metric.

\tableofcontents
\newpage

% =============================================================================
\section{Introduction}\label{sec:intro}
% =============================================================================

The standard model of physics, combined with $\Lambda$CDM cosmology and
general relativity, is phenomenologically successful but treats space as
a pre-existing manifold on which fields propagate. Approximately thirty-five
free parameters govern the combined description, and the origin of the
small cosmological constant, the hierarchy of lepton masses, and the
mechanism of high-temperature superconductivity remain open.

We propose an alternative foundation in which space is not pre-given.
Instead, every source of mass-energy \emph{produces} a local contribution
to a shared scalar field $\Phi$, whose value measures the density of generated
space. The metric, causality and Newtonian gravity emerge as effective
descriptions of the configuration of $\Phi$; the Planck length is the only
invariant carried across scales.

The purpose of this paper is to \emph{define} the theory --- its axioms,
substrate, effective field, and canonical equations --- in a minimal,
self-contained form, and to delineate explicitly what has been proven and
what remains open. We consciously avoid a survey of applications: these are
developed in companion papers that cite this reference (see
Sec.~\ref{sec:applications}).

\paragraph{What TGP is not.} TGP is not a scalar-tensor theory: the
effective metric is algebraically determined by $\Phi$, and the action
contains no Einstein--Hilbert term. TGP is not a programme in which space
emerges from information or entanglement (AdS/CFT, ER\,=\,EPR): here space
emerges from \emph{matter}. TGP is not an extension of general relativity
with additional fields; it is a reformulation in which GR is recovered in
a well-defined regime.

\paragraph{What TGP is.} A single scalar degree of freedom $\Phi$,
coarse-grained from a $\Zp$-symmetric relational substrate, satisfying a
single nonlinear field equation sourced by matter, and inducing a
hyperbolic effective metric whose temporal factor is the normalised
substrate-potential density (vacuum form
$g_{tt}\propto U(\Phi)/\Phi^{4}$). From this minimal structure flow all
predictions of the theory.

\paragraph{Scope of this paper.} Sections~\ref{sec:axioms}--\ref{sec:metric}
establish the axiomatic base, the substrate-to-field derivation, the
effective field theory, and the emergent metric. Section~\ref{sec:regimes}
exhibits the three physical regimes. Section~\ref{sec:status} enumerates the
status of theorems and open problems --- the central artifact of this paper
from a reproducibility standpoint. Section~\ref{sec:applications} is a brief
forward reference to companion papers.

\paragraph{Main open caveat.} The passage from the discrete substrate
Hamiltonian $H_{\Gamma}$ to the continuum TGP action rests on the
Ginzburg--Landau (GL) bond in~\eqref{eq:H-Gamma}, whose continuum form
$\mathcal{F}_{\mathrm{kin}}^{\mathrm{geo}}[\varphi]=(K_{\mathrm{geo}}/2)
\int\varphi^{4}(\nabla\varphi)^{2}\,d^{d}x$ is an elementary Taylor
expansion (App.~B, Prop.~\ref{prop:substrate-action}). What remains
open is the strictly rigorous Banach/$\Gamma$-convergence statement
for the block-spin map; this is a mathematical hardening task, not an
open derivation of $\alpha$. An earlier attempt to derive $\alpha=2$
from a bilinear v1 bond (OP-6 of v1) was shown to fail on three
independent grounds and has been resolved by the axiom pivot
documented in Remark~\ref{rem:alpha2-closure} and
\texttt{KNOWN\_ISSUES.md}. Throughout the paper we present the
continuum field theory of Sec.~\ref{sec:field} as a well-posed
effective theory whose axiomatic continuum limit is explicit.

% =============================================================================
\section{Axioms}\label{sec:axioms}
% =============================================================================

TGP rests on four axioms. Axioms \ref{ax:N0}--\ref{ax:P3} concern the
reference state and its instability; Axiom~\ref{ax:source} states how
matter couples to space. A fifth, derived statement (the $\Zp$ symmetry
of the substrate) appears as a proposition rather than a postulate.

\begin{axiom}[Absolute reference state]\label{ax:N0}
There exists a reference state $\Nzero$ -- \emph{absolute nothingness} --
characterized by the simultaneous vanishing of the state function
$\Ffield$ and the field $\Phi$:
\begin{equation}\label{eq:N0}
  \Nzero := \bigl\{\emptyset,\;\Ffield\bigr\},
  \qquad
  \Ffield\big|_{\Nzero}=0,
  \qquad
  \Phi\big|_{\Nzero}=0.
\end{equation}
In $\Nzero$ there is no metric, no matter, no time, no energy.
$\Nzero$ is \emph{not} the quantum vacuum and \emph{not} empty space;
it is the absence of every physical metric structure.
\end{axiom}

\begin{axiom}[Relational substrate]\label{ax:substrate}
Underlying both the $\Nzero$ state and all physical configurations is a
fundamental, non-emergent relational substrate
\begin{equation}\label{eq:Gamma}
  \Gamma=(V,E),
\end{equation}
with vertices $V$ and edges $E$, possessing adjacency relations and
connectivity but \emph{no a priori metric}. Each node $i\in V$ carries an
amplitude operator $\hat{s}_{i}$. The substrate Hamiltonian is
\begin{equation}\label{eq:H-Gamma}
  H_{\Gamma}=\sum_{i}\!\Bigl[
    \tfrac{\hat{\pi}_{i}^{2}}{2\mu}
    +\tfrac{m_{0}^{2}}{2}\hat{s}_{i}^{2}
    +\tfrac{\lambda_{0}}{4}\hat{s}_{i}^{4}
  \Bigr]
  + J\!\!\sum_{\langle ij\rangle}\! A_{ij}\,
    \hat{s}_{i}^{2}\,\hat{s}_{j}^{2}\,
    \bigl(\hat{s}_{j}^{2}-\hat{s}_{i}^{2}\bigr)^{2},
\end{equation}
with $J>0$, $\lambda_{0}>0$, and $A_{ij}\in\{0,1\}$ the adjacency matrix.
The bond is a Ginzburg--Landau (GL) gradient term in the Z$_{2}$-even
composite $\hat{\Phi}_{i}\equiv\hat{s}_{i}^{2}$, equivalent in continuum
to $\mathcal{F}_{\mathrm{kin}}^{\mathrm{geo}}[\varphi]=
(K_{\mathrm{geo}}/2)\int\varphi^{4}(\nabla\varphi)^{2}\,d^{d}x$ (see
App.~B, Prop.~\ref{prop:substrate-action}); it vanishes identically on
any configuration with $\hat{s}_{i}=0$, encoding the fact that
\emph{an empty substrate cannot support geometry}. This is a change
from v1: see \ref{rem:alpha2-closure} below and
\texttt{KNOWN\_ISSUES.md}.
\end{axiom}

\begin{proposition}[$\Zp$ symmetry]\label{prop:Z2}
The Hamiltonian~\eqref{eq:H-Gamma} is invariant under the chiral inversion
$\hat{s}_{i}\mapsto-\hat{s}_{i}$ for all $i\in V$. This discrete $\Zp$
symmetry is the fundamental symmetry of TGP; the state function $\Ffield$
measures the extent of its spontaneous breaking.
\end{proposition}

\begin{axiom}[Matter generates space]\label{ax:source}
Every source of mass-energy $\rho_{a}$ contributes to the field $\Phi$
according to
\begin{equation}\label{eq:source}
  \DD[\Phi_{a}]\;=\;q_{a}\,\rho_{a},
\end{equation}
where $q_{a}>0$ is the \emph{spatial generation constant} of particle
type $a$ and $\DD$ is the covariant operator governing the dynamics of
$\Phi$ (made explicit in Sec.~\ref{sec:field}).
\end{axiom}

\begin{axiom}[Instability of $\Nzero$]\label{ax:P3}
The state $\Ffield=0$ ($\Nzero$) is unstable: its measure in configuration
space vanishes,
\begin{equation}
  \mu\bigl(\{\sigma\in\Omega : \Ffield[\sigma]=0\}\bigr)=0.
\end{equation}
Any departure from exact $\Zp$ symmetry leads locally to $\Ffield>0$ and
hence to simultaneous matter and generated space.
\end{axiom}

\begin{proposition}[Instability from substrate thermodynamics]\label{prop:instability}
Let $\mu_{\beta}(\sigma)\propto e^{-\beta H_{\Gamma}[\sigma]}$ be the Gibbs
measure for~\eqref{eq:H-Gamma}. In the thermodynamic limit $|V|\to\infty$
there exists $\beta_{c}>0$ such that
$\mu_{\beta}\bigl(\{\Ffield=0\}\bigr)=0$ for all $\beta>\beta_{c}$; the
physical substrate ($\beta\to\infty$) is always in the spontaneously
broken regime. Thus Axiom~\ref{ax:P3} is a \textbf{consequence} of
Axioms~\ref{ax:substrate}--\ref{ax:source} together with standard Ising
thermodynamics, not an independent postulate.
\end{proposition}

\begin{remark}[Three-level ontology]
These axioms generate a three-level hierarchy:
(0) the relational substrate $\Gamma$ is fundamental and non-emergent;
(1) the field $\Phi$ is a coarse-grained \emph{metric-phase parameter} of
the substrate, with $\Phi=0$ the non-metric phase and $\Phi>0$ the metric
phase;
(2) the effective geometry $g_{\mu\nu}$ arises from coarse-graining
level~(1). The transition $\Phi=0\to\Phi>0$ does not create space from
nothing: it switches on the metric response of an already existing
substrate, in the same sense that freezing water does not create
molecules but changes their phase.
\end{remark}

% =============================================================================
\section{From substrate to field}\label{sec:substrate-to-field}
% =============================================================================

The relation between the microscopic substrate $\Gamma$ and the macroscopic
field $\Phi$ is established by coarse-graining. Given a block
$\mathcal{B}(\mathbf{x})$ of lattice sites centred on $\mathbf{x}$,
\begin{equation}\label{eq:Phi-def}
  \Phi(\mathbf{x})
  \;\propto\;
  \langle\hat{s}^{2}\rangle_{\mathcal{B}(\mathbf{x})}.
\end{equation}
In the mean-field approximation one has the identity
\begin{equation}\label{eq:F-s-relation}
  \Ffield_{i}[\Omega]
  \;=\;\langle\Omega|\,\hat{s}_{i}\,|\Omega\rangle^{2},
\end{equation}
so that $\Ffield=0$ iff $\langle\hat{s}_{i}\rangle=0$ (unbroken $\Zp$,
i.e.\ $\Nzero$), while $\Ffield>0$ signals spontaneous breaking and the
simultaneous appearance of matter and space. In the continuum limit,
the bond term in~\eqref{eq:H-Gamma} reduces to the Ginzburg--Landau
gradient functional $\mathcal{F}_{\mathrm{kin}}^{\mathrm{geo}}[\varphi]
=(K_{\mathrm{geo}}/2)\int\varphi^{4}(\nabla\varphi)^{2}d^{d}x$, as made
explicit in App.~B (Prop.~\ref{prop:substrate-action}); the Euler--Lagrange
equation of this functional is the continuum field equation for $\Phi$
that we state and analyse in Sec.~\ref{sec:field}.

\begin{remark}[One substrate, two projections]
Coarse-graining $H_{\Gamma}$ yields \emph{two} emergent fields from a
single microscopic object:
\begin{enumerate}[(i),nosep]
  \item the scalar metric-phase parameter
    $\Phi(\mathbf{x})\propto\langle\hat{s}^{2}\rangle_{\mathcal{B}}$,
    which governs the entire gravitational/cosmological sector;
  \item a spatial-correlation tensor
    $\sigma_{ab}(\mathbf{x})\propto
    \langle\hat{s}_{i}\,\hat{s}_{i+\hat{e}_{a}}\rangle_{\mathcal{B}}^{\mathrm{TF}}$,
    carrying spin-2 and underlying tensor gravitational-wave modes.
\end{enumerate}
Both follow from the \emph{same} $H_{\Gamma}$ and are not independent
postulates. This is crucial: TGP is not a scalar-tensor theory, because
$\Phi$ and $\sigma_{ab}$ are two projections of one substrate rather than
two independent fields.
\end{remark}

% =============================================================================
\section{Effective field theory}\label{sec:field}
% =============================================================================

We use the dimensionless variable $\varphi \equiv \Phi/\PhiZero$.

\subsection{Uniqueness of the kinetic operator}

\begin{theorem}[Uniqueness of $\DD_{\mathrm{kin}}$ within class (C1)--(C3), $\alpha=2$]\label{thm:alpha2}
Among all local second-order scalar kinetic operators of the form
\begin{equation}
  \DD[\varphi]
  = K(\varphi)\,\nabla^{2}\varphi
  + \tfrac{1}{2}K'(\varphi)(\nabla\varphi)^{2},
\end{equation}
imposing
\begin{enumerate}[(C1),nosep]
  \item $K(\varphi)$ smooth and depending only on $\varphi$ (locality);
  \item $\DD[\varphi]\to\nabla^{2}\varphi$ for $\varphi=\mathrm{const}$
    (Laplacian limit);
  \item covariance under conformal rescaling $\varphi\mapsto\lambda\varphi$
    of the substrate metric $h_{ij}=\varphi^{4}\,\delta_{ij}$
    ($\Phi$-covariance),
\end{enumerate}
selects uniquely $K(\varphi)=K_{\mathrm{geo}}\varphi^{4}$ with the
ratio $\alpha(\varphi)\equiv K'(\varphi)\varphi/[2K(\varphi)]=2$ (constant).
The kinetic operator is
\begin{equation}\label{eq:Dkin}
  \boxed{\;
    \DD_{\mathrm{kin}}[\varphi]
    \;=\;\nabla^{2}\varphi+\frac{2(\nabla\varphi)^{2}}{\varphi}
    \;=\;\frac{1}{3\varphi^{2}}\,\nabla^{2}(\varphi^{3}).
  \;}
\end{equation}
\end{theorem}

\begin{proof}[Proof sketch]
Condition (C2) fixes $K(1)=1$. The Euler--Lagrange equation of
$\mathcal{L}_{\mathrm{kin}}=-\tfrac{1}{2}K(\varphi)(\nabla\varphi)^{2}$
yields the ratio $\alpha(\varphi)=K'(\varphi)\varphi/[2K(\varphi)]$.
(C1) forces $\alpha$ to be a constant; (C3) fixes $\alpha=2$ through the
explicit computation of the Laplace--Beltrami operator on
$h_{ij}=\varphi^{4}\,\delta_{ij}$:
$\Delta_{h}\varphi=\varphi^{-4}\,\DD_{\mathrm{kin}}[\varphi]$ holds
iff $\alpha=2$. Hence $K=\varphi^{4}$ (up to the overall constant
$K_{\mathrm{geo}}$, conventionally $1$).
\end{proof}

\begin{remark}[What $\alpha=2$ means; closure of OP-6]\label{rem:alpha2-closure}
The exponent $\alpha=2$ is \emph{not} fitted to data: it is fixed by the
geometric coupling $K_{ij}=J(\varphi_{i}\varphi_{j})^{2}$ between nearest
neighbours on the substrate, under the three conditions (C1)--(C3). The
statement is therefore a uniqueness result \emph{within} the stated
class of local, $\Phi$-covariant second-order scalar operators; relaxing
any of (C1)--(C3) (non-locality, higher derivatives, or departures from
the conformal rescaling of the substrate metric) would open additional
channels that are not considered by TGP.

In v1 of this paper (Zenodo DOI \texttt{10.5281/zenodo.19670324}) the
substrate Hamiltonian carried a linear bond $-J\hat{s}_{i}\hat{s}_{j}$;
the statement that the continuum limit generates the GL coupling
$K_{ij}=J(\varphi_{i}\varphi_{j})^{2}$, and thus supplies (C1)--(C3)
from below, was listed as open problem OP-6. Between 2026-04-23 and
2026-04-24, OP-6 was investigated on three independent grounds:
(a)~the analytical sketch of standard RG mechanisms (M2-a); (b)~1D
Monte Carlo of the envelope stiffness and of the block-RG bond flow
(M2-b, M3-a); (c)~scaling-dimension analysis at the 3D Ising
Wilson--Fisher fixed point using conformal-bootstrap data (M3-c). All
three rule out the v1 derivation: at 3D Ising WF, the ratio
$K^{(1)}/K^{(0)}$ decays in the IR as $k^{\Delta_{\epsilon}}=k^{1.41}$,
so $K_{*}(\varphi)\to\mathrm{const}$ rather than $K_{*}(\varphi)
\propto\varphi$. Numerically, the 1D block-RG measurement returns
$K_{B}(\langle\Phi_{B}\rangle)\propto\langle\Phi_{B}\rangle^{-0.52\pm0.28}$
uniformly across block sizes $B\in\{1,2,4,8,16\}$, rejecting the
target $p_{\Phi}=+1$ at $5.5\sigma$.

The present paper (v2 in preparation) therefore adopts the GL bond
directly as the substrate axiom~\eqref{eq:H-Gamma}. Under this axiom,
(C1)--(C3) and the GL gradient functional $\mathcal{F}_{\mathrm{kin}}^{\mathrm{geo}}$
of Prop.~\ref{prop:substrate-action} (App.~B) are \emph{immediate},
and Theorem~\ref{thm:alpha2} reduces to the statement that any
axiomatic choice respecting substrate--metric conformal covariance
must be of this GL form. OP-6 is closed (see
\texttt{TGP\_v1/research/op6/} for the complete record). The bond
vanishes on empty-substrate configurations ($\hat{s}_{i}=0$),
making ``no substrate~$\Rightarrow$ no geometry'' a property of the
axiom rather than a derived fact.
\end{remark}

\subsection{Action, potential, and field equation}

The complete TGP action is
\begin{equation}\label{eq:S-TGP}
  S_{\mathrm{TGP}}[\Phi,\psi_{m}]
  =\int d^{4}x\,\sqrt{-g_{\mathrm{eff}}}\,
    \Bigl[\mathcal{L}_{\mathrm{field}}(\Phi)
    +\mathcal{L}_{\mathrm{mat}}(\Phi,\psi_{m})\Bigr],
\end{equation}
with field Lagrangian
\begin{equation}\label{eq:L-field}
  \mathcal{L}_{\mathrm{field}}
  =-\tfrac{1}{2}\,K(\varphi)\,
    g_{\mathrm{eff}}^{\mu\nu}\partial_{\mu}\varphi\,\partial_{\nu}\varphi
  - U(\varphi),
  \qquad K(\varphi)=\varphi^{4},
\end{equation}
self-interference potential
\begin{equation}\label{eq:U-potential}
  U(\varphi)
  =\frac{\beta}{3}\varphi^{3}-\frac{\gamma}{4}\varphi^{4},
  \qquad \beta=\gamma
  \;\;\text{(vacuum condition $U'(1)=0$)},
\end{equation}
and minimal matter coupling
$\mathcal{L}_{\mathrm{mat}}=-\,q\,\varphi\rho/\PhiZero$. The volume element
is $\sqrt{-g_{\mathrm{eff}}}=c_{0}\varphi$ (see Sec.~\ref{sec:metric}).

\begin{theorem}[Full TGP field equation]\label{thm:field-eq}
Variation $\delta S_{\mathrm{TGP}}/\delta\Phi=0$ of~\eqref{eq:S-TGP} yields
\begin{equation}\label{eq:field-eq}
  \boxed{\;
    \nabla^{2}\Phi
    +\frac{2(\nabla\Phi)^{2}}{\Phi}
    +\beta\,\frac{\Phi^{2}}{\PhiZero}
    -\gamma\,\frac{\Phi^{3}}{\PhiZero^{2}}
    \;=\;-\,q\,\PhiZero\,\rho.
  \;}
\end{equation}
\end{theorem}

\begin{remark}[TGP is not a scalar-tensor theory]\label{rem:not-ST}
The action~\eqref{eq:S-TGP} contains \emph{no} Einstein--Hilbert term $R$.
The effective metric $g_{\mu\nu}^{\mathrm{eff}}$ is defined algebraically
by $\Phi$ (not an independent dynamical variable). The Einstein equations
arise as a geometric consistency condition of $g_{\mu\nu}^{\mathrm{eff}}$,
propagated by the Bianchi identity once satisfied at an initial time; they
are not postulated. The only dynamical degree of freedom is $\Phi$.
\end{remark}

\begin{remark}[Formal contrast with scalar-tensor theories]\label{rem:ST-formal}
Structurally TGP differs from the Brans--Dicke / Horndeski family on three
counts, each visible at the level of the action and the equations of
motion:
\begin{enumerate}[(i),nosep]
  \item \emph{Independent tensor d.o.f.\ absent.} In scalar-tensor
    theories the action is $S=\int\sqrt{-g}\,[f(\phi)R/16\pi G+\ldots]$
    with $g_{\mu\nu}$ varied independently. Here the only field varied
    is $\Phi$; the metric is algebraically determined by
    $g_{\mu\nu}^{\mathrm{eff}}[\Phi]$ through Eq.~\eqref{eq:exp-metric}.
  \item \emph{No conformal frame ambiguity.} Scalar-tensor theories
    admit Jordan/Einstein frame reparametrisations. TGP has a single
    frame: the substrate metric $h_{ij}=\Phi^{4}\delta_{ij}$ is fixed by
    the $\Zp$ substrate structure; no conformal redefinition of the
    ``physical'' metric is available.
  \item \emph{Kinetic operator not a free function.} In scalar-tensor
    theories $f(\phi)$ is an input function to be fitted. Here
    $K(\varphi)=\varphi^{4}$ is fixed by
    Theorem~\ref{thm:alpha2} under (C1)--(C3), not chosen by hand.
\end{enumerate}
These three properties preclude TGP from being recast as a standard
scalar-tensor theory in any frame; the GR limit is recovered only as an
effective low-curvature description, not as a conformal transformation
of a two-metric theory.
\end{remark}

\subsection{Ghost-freedom}\label{ssec:ghost}

The substrate coupling $K_{\mathrm{sub}}(\varphi)=\varphi^{4}$ is
strictly positive for all $\varphi>0$, so the kinetic term
$\tfrac{1}{2}\varphi^{4}(\nabla\varphi)^{2}\geq 0$ is positive-semidefinite.
TGP is ghost-free at three levels simultaneously:

\begin{center}\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Sector} & \textbf{Kinetic positivity} & \textbf{Mechanism}\\
\midrule
Fundamental ($\varphi$) & $\varphi^{4}(\nabla\varphi)^{2}>0$ & geometric\\
Mukhanov--Sasaki perturbations & $Q_{s}=\varphi_{\mathrm{bg}}^{4}>0$ & linearized\\
Solitons (full lattice) & $K_{\mathrm{sub}}=\varphi^{4}>0$ & substrate\\
\bottomrule
\end{tabular}
\end{center}

A logarithmic truncation $K=1+4\ln\varphi$ that appears in certain
expansions exhibits an apparent zero at $\varphi^{\ast}=e^{-1/4}$; this
is an artifact of the Taylor expansion of the exponent,
$\varphi^{4}=e^{4\ln\varphi}=1+4\ln\varphi+\dots$, not a physical
singularity.

% =============================================================================
\section{Emergent metric}\label{sec:metric}
% =============================================================================

The effective metric of TGP is not postulated. It is derived from three
inputs:
(i) substrate-budget condition $\Rightarrow f\cdot h=1$ (antipodal projection);
(ii) substrate-potential density $\Rightarrow f(\psi)\propto U(\Phi)/\Phi^{4}$
(normalised to vacuum);
(iii) full $\Phi$-EOM consistency at the post-Newtonian level
$\Rightarrow \gamma_{\mathrm{PPN}}=\beta_{\mathrm{PPN}}=1$ at 1PN with
explicit, falsifiable 2PN deviations.

\subsection{Spatial part: node-density ansatz}

A coarse-graining cell $\mathcal{B}(\mathbf{x})$ containing $N_{B}$
lattice nodes has physical volume proportional to $\Phi(\mathbf{x})$:
\begin{equation}\label{eq:g-spatial}
  g_{ij}=h(\psi)\,\delta_{ij},
  \qquad \psi\equiv\Phi/\PhiZero,
\end{equation}
with $h(\psi)$ fixed below by the substrate budget and
the substrate-potential density.

\subsection{Substrate budget: $f\cdot h=1$}

The substrate has a fixed information capacity $\mathcal{B}=N_{B}s_{0}$ per
cell. Partitioning between spatial capacity $h(\Phi)$ and temporal
capacity $f(\Phi)$, the multiplicative conservation law
\begin{equation}\label{eq:fh-1}
  f(\Phi)\cdot h(\Phi)=1
\end{equation}
holds across all scales. This single condition is sufficient to fix
$\gamma_{\mathrm{PPN}}=1$ algebraically: in the weak-field expansion
$f\to 1+f_{1}U+\ldots$ with the budget enforcing $h=1/f$, the spatial
expansion of any test geodesic acquires the same coefficient as the
temporal expansion, so $\gamma_{\mathrm{PPN}}\equiv 1$ regardless of
the explicit form of $f$.

\subsection{Hyperbolic metric from substrate-potential density}

\begin{theorem}[Substrate-potential metric ansatz, within the TGP
ansatz class $fh=1$]\label{thm:metric}
Within the TGP ansatz class enforcing the substrate budget
$f\cdot h=1$ (Eq.~\eqref{eq:fh-1}), the temporal factor $f$ is the
normalised substrate-potential density:
\begin{equation}\label{eq:metric-from-V}
  f(\psi)
  \;=\;
  \frac{12}{\gamma\,\PhiZero^{2}}\,\frac{U(\Phi)}{\psi^{4}}
  \;=\;
  \frac{4-3\psi}{\psi},
  \qquad
  h(\psi)=\frac{1}{f(\psi)}=\frac{\psi}{4-3\psi},
\end{equation}
where $U(\Phi)$ is the substrate self-interaction potential
of Eq.~\eqref{eq:U-potential} with $\beta=\gamma$ (vacuum condition).
The boxed effective line element is
\begin{equation}\label{eq:exp-metric}
  \boxed{\;
    ds^{2}
    =-c_{0}^{2}\,\frac{4-3\psi}{\psi}\,dt^{2}
    +\frac{\psi}{4-3\psi}\,\delta_{ij}\,dx^{i}dx^{j},
    \qquad
    \psi=\Phi/\PhiZero.
  \;}
\end{equation}
At the vacuum minimum $\psi=1$ one has $f=h=1$ (calibration); at
$\psi=4/3$ the temporal factor vanishes (boundary of the ghost-free
basin, Sec.~\ref{ssec:ghost}); at $\psi\to 0$ the temporal factor
diverges (non-metric phase). The hyperbolic ansatz, combined with the
full nonlinear $\Phi$-EOM~\eqref{eq:field-eq} with $\alpha=2$, yields
$\gamma_{\mathrm{PPN}}=\beta_{\mathrm{PPN}}=1$ \emph{exactly at 1PN}
(Cor.~\ref{cor:ppn}); deviations from GR appear explicitly at the
2PN level and are falsifiable predictions of TGP.
\end{theorem}

\begin{remark}[Three independent substrate principles converge on
$f=(4-3\psi)/\psi$]\label{rem:metric-triple}
The hyperbolic form~\eqref{eq:metric-from-V} is uniquely picked by
\emph{three} logically independent substrate principles
(audit: \texttt{KNOWN\_ISSUES.md} 2026-04-25 M9.1$''$ P2):
(P2-C) $f$ is the rational function of minimal degree compatible with
the budget $fh=1$, the vacuum calibration $f(1)=1$, and the boundary
condition $f(4/3)=0$ at the second zero of $U(\Phi)$
(\texttt{prop:ghost-free}); (P2-D) $f$ is the lowest-derivative
dimensionless ratio $U(\Phi)/\Phi^{4}$ (in natural units
$[\Phi]=\mathrm{mass}$, $[U]=\mathrm{mass}^{4}$); (P2-E) $f-1$
correlates with the substrate energy density excess above vacuum.
Triple convergence eliminates arbitrariness; in TGP's emergent-gravity
philosophy the hyperbolic form is the natural substitute for an
otherwise unavailable single-action variation, since the metric is not
an autonomous field varied independently of $\Phi$
(Rem.~\ref{rem:not-ST}).
\end{remark}

\begin{remark}[Scope of the selection statement; closure of the
power-law branch]\label{rem:metric-scope}
Theorem~\ref{thm:metric} is a selection statement \emph{within} the TGP
ansatz class $fh=1$ together with the substrate-potential density
postulate~\eqref{eq:metric-from-V}. It is not a uniqueness statement
against every conceivable $(f,h)$; a parametrically distinct metric
structure (e.g.\ bi-metric or tensor-valued) would not be in
competition here. The earlier power-law family $h=\psi^{p}$ explored
in v1 of this paper has been closed \emph{negatively} by a master
formula $\beta_{\mathrm{PPN}}(p,\alpha)=(p-1-\alpha)/p$ that admits no
admissible solution with $\alpha>0$
(\texttt{KNOWN\_ISSUES.md} 2026-04-25 M9.1, M9.1$'$). The
substrate-potential ansatz~\eqref{eq:metric-from-V} is the unique
pivot consistent with current post-Newtonian data within the $fh=1$
class.
\end{remark}

\begin{corollary}[PPN parameters at 1PN, within the ansatz of
Theorem~\ref{thm:metric}]\label{cor:ppn}
For the hyperbolic metric ansatz~\eqref{eq:exp-metric} solved together
with the full nonlinear $\Phi$-EOM~\eqref{eq:field-eq} with $\alpha=2$,
the post-Newtonian expansion yields
\begin{equation}\label{eq:ppn}
  \gamma_{\mathrm{PPN}}=1,\qquad \beta_{\mathrm{PPN}}=1
  \qquad\text{(exact at 1PN)}.
\end{equation}
The expansion of $g_{tt}$ in the Newtonian potential
$U_{\mathrm{N}}=GM/(c_{0}^{2}r)$ continues with explicit deviations
from the Schwarzschild isotropic form beginning at 2PN:
\begin{align}\label{eq:gtt-expansion}
  g_{tt}^{\mathrm{TGP}}/(-c_{0}^{2})
  &=1-2U_{\mathrm{N}}+2U_{\mathrm{N}}^{2}-\tfrac{7}{3}U_{\mathrm{N}}^{3}
   +\tfrac{35}{12}U_{\mathrm{N}}^{4}+\mathcal{O}(U_{\mathrm{N}}^{5}),
   \nonumber\\
  g_{tt}^{\mathrm{Schw}}/(-c_{0}^{2})
  &=1-2U_{\mathrm{N}}+2U_{\mathrm{N}}^{2}-\tfrac{3}{2}U_{\mathrm{N}}^{3}
   +U_{\mathrm{N}}^{4}-\ldots,
\end{align}
giving a 2PN fractional deviation
$|\Delta g_{tt}|_{2\mathrm{PN}}=\tfrac{5}{6}\,U_{\mathrm{N}}^{3}$ as
\emph{the} concrete falsifiable prediction of TGP at the post-Newtonian
level (\texttt{KNOWN\_ISSUES.md} 2026-04-25 M9.1$''$ P1).
Current empirical bounds at 1PN are passed exactly:
Cassini Shapiro-delay tracking gives
$|\gamma_{\mathrm{PPN}}-1|\leq 2.3\times 10^{-5}$~\cite{BertottiCassini},
and Lunar Laser Ranging (Nordtvedt effect) gives
$|\beta_{\mathrm{PPN}}-1|\leq 6\times 10^{-5}$~\cite{WilliamsLLR}.
At 2PN, the predicted deviation is far below current solar-system
sensitivity ($U_{\mathrm{N}}\sim 10^{-8}\Rightarrow|\Delta|\sim 10^{-23}$);
binary pulsar timing
($U_{\mathrm{N}}\sim 10^{-6}\Rightarrow|\Delta|\sim 10^{-17}$) is
likewise unconstraining. The most decisive future constraints come
from GW170817-class binary inspirals (conditional on tensor-sector
closure, OP-7) and EHT photon rings (strong field,
$U_{\mathrm{N}}\sim 1/3$); see \texttt{KNOWN\_ISSUES.md} 2026-04-25
M9.1$''$ P3 for the audit (3 PASS, 1 conditional tension, 1 open,
0 falsified).
\end{corollary}

\begin{corollary}[Luminal gravitational waves -- conditional on
tensor sector OP-7]\label{cor:cGW}
For the scalar fluctuation sector of $\Phi$, fluctuations propagate on
the effective metric $g_{\mathrm{eff}}$ at the vacuum speed
$c_{\mathrm{GW}}=c_{0}$ identically, consistent with GW170817
constraints ($|c_{\mathrm{GW}}-c_{0}|/c_{0}<10^{-15}$). The full
prediction for the two transverse-traceless tensor polarisations of
gravitational waves requires the tensor sector OP-7 (open problem;
\texttt{KNOWN\_ISSUES.md} C4); the scalar-sector luminality is the
piece currently established.
\end{corollary}

\subsection{Field-dependent constants}

Requiring invariance of the Planck length $\lP=\sqrt{G\hbar/c^{3}}$
under the emergent rescaling fixes the field dependence of the
fundamental constants:
\begin{equation}\label{eq:constants}
  c(\Phi)=\frac{c_{0}}{\sqrt{\psi}},
  \qquad
  G(\Phi)=\frac{G_{0}}{\psi},
  \qquad
  \hbar(\Phi)=\frac{\hbar_{0}}{\sqrt{\psi}},
\end{equation}
all consistent with the hyperbolic metric~\eqref{eq:exp-metric}. $\lP$
is the sole invariant across scales.

% =============================================================================
\section{Three physical regimes}\label{sec:regimes}
% =============================================================================

The non-linear potential $U(\varphi)$ together with the source term in
Eq.~\eqref{eq:field-eq} generates three distinct interaction regimes from a
single field equation. Linearization around a single static point source
yields a radial profile $\phi(r)$ with the following properties:

\begin{center}\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Regime} & \textbf{Scale} & \textbf{Physical content}\\
\midrule
(i)\ long-range attraction   & $r\gg r_{\mathrm{rep}}$ & Newtonian gravity, $U\sim -GM/r$\\
(ii)\ intermediate repulsion & $r\sim r_{\mathrm{rep}}$ & positive spatial potential between sources\\
(iii)\ short-range well      & $r\lesssim r_{\mathrm{well}}$ & confinement ($U\to-\infty$ finite well)\\
\bottomrule
\end{tabular}
\end{center}

The existence of three regimes follows from the non-monotonicity of
$\phi(r)$: a single generation profile acquires at least one inflection
point from the interplay of $\DD_{\mathrm{kin}}[\varphi]$ with the cubic
and quartic terms of $U(\varphi)$. The explicit scales $r_{\mathrm{rep}}$
and $r_{\mathrm{well}}$ depend on the source strength $qM/\PhiZero$ and on
the coupling $\gamma$ through
\begin{equation}\label{eq:Lambda-E}
  \Lambda_{E}\equiv\gamma/12,
  \qquad
  3H_{0}^{2}\approx c_{0}^{2}\,\Lambda_{E}
  \quad(\text{de Sitter limit}),
\end{equation}
which identifies $\Lambda_{E}$ as the effective cosmological constant.

\begin{remark}[Gravity as regime (i)]
Newtonian gravity is recovered as the long-range limit of the full
non-linear equation~\eqref{eq:field-eq}: in the weak field
$\Phi\approx\PhiZero(1+2U/c_{0}^{2})$, the leading source term reproduces
Poisson's equation $\nabla^{2}U=4\pi G\rho$ with
$G=q\PhiZero c_{0}^{2}/(8\pi)$ (identifying $q$, $\PhiZero$ with
Newton's constant).
\end{remark}

% =============================================================================
\section{Status of theorems and open problems}\label{sec:status}
% =============================================================================

We distinguish three levels of status, as follows:
\textsf{[TH]} formally proven theorem;
\textsf{[NU]} numerically verified to stated precision;
\textsf{[HY]} conjecture or hypothesis, not yet proven.
The list below summarises the current (April 2026) status of TGP core.

\subsection{Proven results (Core)}

\begin{center}\small
\begin{longtable}{@{}clp{7.2cm}c@{}}
\toprule
\# & \textbf{Result} & \textbf{Content} & \textbf{Status}\\
\midrule
\endhead
1 & Uniqueness of $\DD_{\mathrm{kin}}$
  & Theorem~\ref{thm:alpha2}: $K(\varphi)=\varphi^{4}$, $\alpha=2$ unique under (C1)--(C3)
  & \textsf{TH}\\
2 & Ghost-freedom (three sectors)
  & Sec.~\ref{ssec:ghost}: $K_{\mathrm{sub}}>0$ $\forall\varphi>0$; no kinetic singularity
  & \textsf{TH}\\
3 & Full field equation
  & Theorem~\ref{thm:field-eq}: Eq.~\eqref{eq:field-eq} from $\delta S/\delta\Phi=0$
  & \textsf{TH}\\
4 & Substrate-potential metric ansatz
  & Theorem~\ref{thm:metric}: $g_{tt}=-c_{0}^{2}(4-3\psi)/\psi$ from $fh=1$ + substrate-potential density;
    triple convergence (P2-C/D/E)
  & \textsf{TH}\\
5 & PPN parameters at 1PN
  & Corollary~\ref{cor:ppn}: $\gamma_{\mathrm{PPN}}=\beta_{\mathrm{PPN}}=1$ exactly at 1PN
    (M9.1$''$ P1 verified); 2PN deviation $|\Delta|=\tfrac{5}{6}U_{\mathrm{N}}^{3}$ predicted
  & \textsf{TH}\\
6 & Luminal scalar-sector GW (conditional)
  & Corollary~\ref{cor:cGW}: scalar sector $c_{\mathrm{GW}}=c_{0}$ identically;
    full tensor sector pending OP-7
  & \textsf{TH}\\
7 & Invariance of $\lP$
  & Eq.~\eqref{eq:constants}: $\lP=\sqrt{G\hbar/c^{3}}=\mathrm{const}$ consistent with $g_{\mu\nu}^{\mathrm{eff}}$
  & \textsf{TH}\\
8 & Instability of $\Nzero$
  & Proposition~\ref{prop:instability}: from Ising thermodynamics of $H_{\Gamma}$
  & \textsf{TH}\\
9 & $\Zp$ substrate symmetry
  & Proposition~\ref{prop:Z2}: $H_{\Gamma}$ invariant under $\hat{s}_{i}\mapsto-\hat{s}_{i}$
  & \textsf{TH}\\
10 & Three regimes from $U(\varphi)$
  & Sec.~\ref{sec:regimes}: non-monotonic $\phi(r)$ from cubic+quartic $U$
  & \textsf{TH}\\
11 & Bianchi propagation of Friedmann constraint
  & Remark following Eq.~(\ref{eq:field-eq}): $\nabla_{\mu}G_{\mathrm{eff}}^{\mu\nu}=0$
  & \textsf{TH}\\
12 & Dark-energy de~Sitter limit
  & Eq.~\eqref{eq:Lambda-E}: $3H_{0}^{2}\approx c_{0}^{2}\Lambda_{E}$, $\Lambda_{E}=\gamma/12$
  & \textsf{TH}\\
13 & Matter-source covariance
  & Axiom~\ref{ax:source} implemented by $\mathcal{L}_{\mathrm{mat}}=-q\varphi\rho/\PhiZero$
  & \textsf{TH}\\
14 & Vacuum condition $\beta=\gamma$
  & From $U'(1)=0$; Eq.~\eqref{eq:U-potential}
  & \textsf{TH}\\
15 & MS kinetic positivity in FRW
  & $Q_{s}=\varphi_{\mathrm{bg}}^{4}>0$ for all cosmological backgrounds
  & \textsf{TH}\\
\bottomrule
\end{longtable}
\end{center}

\subsection{Numerically verified (to stated precision)}

\begin{center}\small
\begin{longtable}{@{}clp{7.2cm}c@{}}
\toprule
\# & \textbf{Result} & \textbf{Content} & \textbf{Status}\\
\midrule
\endhead
16 & FRG structural stability of $K(\rho)\!\propto\!\rho$
  & One-loop LPA' flow near the Wilson--Fisher fixed point in $d=3$,
    $Z_{2}$ class: the kinetic form $K(\rho)\!\propto\!\rho$ is
    preserved, $K_{\mathrm{IR}}/K_{\mathrm{UV}}=1.000$, $\eta^{*}=0.044$.
    This verifies kinetic stability of the axiomatic GL form
    of~\eqref{eq:H-Gamma} at the continuum level; $\alpha=2$ itself
    is supplied by the axiom (see Remark~\ref{rem:alpha2-closure}).
  & \textsf{NU}\\
17 & Cosmology: $\Delta\chi^{2}=+1.8$, $\Delta\mathrm{AIC}=+3.8$
  & vs.\ $\Lambda$CDM on Planck\,+\,DESI data, Friedmann~\eqref{eq:field-eq}
  & \textsf{NU}\\
18 & Solitonic $N$-body stability
  & Chaos suppression, Earnshaw violation; 10/10 tests pass
  & \textsf{NU}\\
19 & Metric derivation chain
  & Spatial $\to$ information $\to$ PPN; cross-checked with $\lP=\mathrm{const}$
  & \textsf{NU}\\
20 & Three-regimes existence
  & Numerical profile $\phi(r)$ with attraction, repulsion, confinement
  & \textsf{NU}\\
21 & Coarse-graining $H_{\Gamma}\!\to\!S_{\mathrm{TGP}}$ (axiomatic + numerical)
  & In the v2 axiom, the GL bond of~\eqref{eq:H-Gamma} is by
    construction the lattice version of
    $\mathcal{F}_{\mathrm{kin}}^{\mathrm{geo}}[\varphi]=(K_{\mathrm{geo}}/2)
    \int\varphi^{4}(\nabla\varphi)^{2}d^{d}x$ (App.~B,
    Prop.~\ref{prop:substrate-action}); no numerical extraction of
    $\alpha$ is required. Numerical block-spin Monte Carlo studies
    that originally probed the v1 bilinear bond
    (\texttt{research/continuum\_limit/cg\_strong\_numerical.py},
    \texttt{TGP\_v1/research/op6/m3a\_block\_rg\_1d.py}) are retained
    as supporting evidence for the axiom-level pivot; the complete
    record is in \texttt{KNOWN\_ISSUES.md} (2026-04-24) and in
    \texttt{TGP\_v1/research/op6/}.
  & \textsf{NU}\\
\bottomrule
\end{longtable}
\end{center}

\subsection{Open problems}

The following are explicitly open: they define the research frontier of
TGP at the core level.

\begin{center}\small
\begin{longtable}{@{}clp{8.5cm}c@{}}
\toprule
\# & \textbf{Open problem} & \textbf{Statement} & \textbf{Status}\\
\midrule
\endhead
OP-1 & Uniqueness of $U(\varphi)$
  & Is the cubic+quartic form the unique self-interference potential
  compatible with $\Zp$ symmetry and the vacuum condition?
  & \textsf{HY}\\
OP-2 & Parametric reduction $\beta=\gamma$
  & Formal proof (beyond the vacuum condition) that $\beta=\gamma$ is
  protected at all scales.
  & \textsf{HY}\\
OP-3 & Identification $a_{\Gamma}=1/\PhiZero$
  & Whether the two fitted parameters coincide, reducing the free input
  count to one.
  & \textsf{HY}\\
OP-4 & Derivation of the quartic coefficient $\gamma$
  & An \textit{a priori} determination of $\gamma$ (equivalently
  $\Lambda_{E}$) from $H_{\Gamma}$ alone.
  & \textsf{HY}\\
OP-5 & Full non-linear $N$-body regime
  & Existence/uniqueness of solutions for strongly overlapping sources
  beyond perturbation theory.
  & \textsf{HY}\\
OP-6 & Substrate metric continuum limit (rigorous)
  & Rigorous Banach contraction of the block-spin map and
  $\Gamma$-convergence of the energy functional from the v2 GL
  bond~\eqref{eq:H-Gamma} to the continuum
  action~\eqref{eq:S-TGP}. \textbf{Axiomatic closure: done}
  (2026-04-24, axiom pivot; see Remark~\ref{rem:alpha2-closure}
  and \texttt{KNOWN\_ISSUES.md}). Mathematical hardening
  (strict Banach/$\Gamma$-convergence theorem): remains open.
  & \textsf{HY}\\
OP-7 & Tensor projection $\sigma_{ab}$
  & Derivation of the explicit tensor-field equation from the same
  $H_{\Gamma}$; GW polarisation content.
  & \textsf{HY}\\
OP-8 & Gauge-sector emergence
  & Emergence of U(1)$\times$SU(2)$\times$SU(3) from the substrate
  (addressed partially in the extended manuscript).
  & \textsf{HY}\\
OP-9 & Strong-field black-hole solutions
  & Complete catalogue of non-singular BH solutions; quasi-normal modes.
  & \textsf{HY}\\
OP-10 & Quantum theory of $\Phi$
  & Second quantisation of $\Phi$ beyond the MS linear regime.
  & \textsf{HY}\\
\bottomrule
\end{longtable}
\end{center}

% =============================================================================
\section{Applications and falsifiability}\label{sec:applications}
% =============================================================================

The minimal core of TGP defined above motivates several companion
studies in which the same substrate and hyperbolic
substrate-potential metric~\eqref{eq:exp-metric} are
confronted with separate empirical domains. Each companion paper cites
this reference for its foundational content; the items below are
pointers, not self-contained derivations.

\begin{itemize}[nosep]
  \item \textbf{Superconductivity.} Using the substrate harmonic
    $a^{\ast}_{\mathrm{TGP}}=7.725\,a_{\mathrm{B}}\approx 4.088$\,\AA\
    together with the orbital-amplitude spectrum
    $\{A_{s},A_{sp},A_{d},A_{f}\}$, the companion paper \cite{TGP-SC}
    presents a single-formula fit of $T_{c}$ across cuprates,
    phonon-mediated metals, pressure-driven systems and lanthanide
    hydrides in terms of three universal TGP constants plus tabulated
    atomic inputs. Scope, systematics and excluded families are
    discussed there.
  \item \textbf{Lepton masses.} A single substrate parameter
    $g_{0}^{e}=0.86941$ reproduces $m_{\mu}/m_{e}$ and $m_{\tau}/m_{e}$
    to better than $10^{-3}$, and the Koide combination $K=2/3$ is
    recovered as an arithmetic identity of the Brannen parametrisation
    \cite{TGP-leptons}; the identification of $g_{0}^{e}$ with a
    substrate quantity is a working hypothesis of that companion paper.
  \item \textbf{Black-hole shadows.} In the strong-field regime the
    hyperbolic substrate-potential metric~\eqref{eq:exp-metric} leads to
    photon-ring properties non-perturbatively distinct from
    Schwarzschild; the post-Newtonian expansion that fixes
    Eq.~\eqref{eq:gtt-expansion} no longer converges at
    $U_{\mathrm{N}}\sim 1/3$, so a quantitative EHT prediction
    requires a dedicated nonlinear analysis (open programme OP-EHT;
    see \texttt{KNOWN\_ISSUES.md} 2026-04-25 M9.1$''$ P3) together
    with the tensor sector of OP-7. Ref.~\cite{TGP-BH} reports the
    weak-field photon-ring estimate from the v1 exponential ansatz;
    re-evaluation under~\eqref{eq:exp-metric} is the natural next
    step.
  \item \textbf{$N$-body dynamics.} Numerical integration of
    Eq.~\eqref{eq:field-eq} in configurations dominated by the
    intermediate-repulsion regime (regime~ii) exhibits chaos suppression
    and a violation of Earnshaw's theorem; see \cite{TGP-nbody} for the
    test suite.
\end{itemize}

\paragraph{Minimal falsification targets.} The following predictions are
of a binary character and provide the most direct falsification tests of
the core theory:
\begin{enumerate}[(F1),nosep]
  \item $\alpha=2$ \emph{identically} (Theorem~\ref{thm:alpha2}):
    any observable inconsistent with $K\propto\varphi^{4}$ at the
    substrate level falsifies TGP.
  \item $\gamma_{\mathrm{PPN}}=\beta_{\mathrm{PPN}}=1$ \emph{at 1PN
    exactly, with explicit 2PN deviations}
    (Corollary~\ref{cor:ppn}): the unique falsifiable signature of TGP
    at the post-Newtonian level is
    $|\Delta g_{tt}|_{2\mathrm{PN}}=\tfrac{5}{6}\,U_{\mathrm{N}}^{3}$
    (Eq.~\eqref{eq:gtt-expansion}); a deviation from this prediction at
    GW170817-class binary inspirals (conditional on OP-7) or at EHT
    photon rings (strong field, OP-EHT) would falsify the
    substrate-potential ansatz.
  \item $c_{\mathrm{GW}}=c_{0}$ \emph{identically in the scalar
    sector} (Corollary~\ref{cor:cGW}): already supported by GW170817 at
    the $10^{-15}$ level. The tensor sector is open (OP-7).
\end{enumerate}

\paragraph{Closed-branch negative checks.} Three potential near-term
falsification channels have been analysed in dedicated companion
studies and found to lie \emph{below} current experimental precision,
closing them as viable discovery avenues without contradicting the
theory:
\begin{itemize}[nosep]
  \item \emph{Muon $(g\!-\!2)$:} the TGP-substrate correction to the
    anomalous magnetic moment is $\Delta a_{\mu}\lesssim 10^{-12}$,
    three orders of magnitude below the current experimental
    uncertainty; see \texttt{research/muon\_g\_minus\_2/}.
  \item \emph{Wiedemann--Franz in bad metals:} the substrate-diffusion
    correction to $L/L_{0}$ is $\mathcal{O}(10^{-4})$ at $300$\,K, three
    to four orders of magnitude smaller than observed deviations in
    cuprates and heavy-fermion compounds, which are therefore \emph{not}
    of TGP origin; see \texttt{research/thermal\_transport\_molecular/}.
  \item \emph{Casimir force in MOFs:} the TGP correction is at the
    $10^{-4}$--$10^{-3}$ level, below current torsion-balance precision;
    see \texttt{research/casimir\_mof/}.
\end{itemize}
These null results are stated not as corroborations but as explicit
constraints on where a TGP-specific signature could plausibly emerge
with present instrumentation.

% =============================================================================
\section{Conclusion}\label{sec:conclusion}
% =============================================================================

We have presented a minimal, self-contained core of the Theory of Generated
Space: four axioms, one scalar field with a geometrically fixed kinetic
operator ($\alpha=2$), a single non-linear potential with two coefficients
related by a vacuum condition, and an algebraically determined effective
metric whose temporal factor is the normalised substrate-potential density
$U(\Phi)/\Phi^{4}$, reproducing general relativity exactly at 1PN with
explicit, falsifiable 2PN-level deviations. Fifteen core results are
formally proven, six more are numerically verified, and ten explicit open
problems are stated so that the theory is both well-defined and falsifiable
on a clear basis.

This core is intended to serve as a common reference for applications to
superconductivity, particle masses, black-hole phenomenology and cosmology,
developed in companion papers.

\paragraph{Data and reproducibility.} All numerical scripts underlying the
results of this paper are available in the accompanying source repository
\cite{TGP-repo}. The present paper deposits a fixed, versioned snapshot on
Zenodo (DOI assigned at publication).

\paragraph{Acknowledgements.} The author thanks the community of
open-source physics for the tools that made this work possible.

% =============================================================================
\begin{thebibliography}{99}
% =============================================================================

\bibitem{TGP-repo}
M.~Serafin, \emph{Theory of Generated Space --- Core paper source repository},
\url{https://github.com/Stefan13610/tgp-core-paper} (2026).

\bibitem{TGP-SC}
M.~Serafin,
\emph{Unified superconductivity from the TGP substrate},
companion paper, Zenodo (2026).

\bibitem{TGP-leptons}
M.~Serafin,
\emph{Lepton masses and the Koide relation from a single TGP parameter},
companion paper (2026).

\bibitem{TGP-BH}
M.~Serafin,
\emph{Black-hole shadows in TGP: weak-field estimate (v1 exponential
ansatz); strong-field analysis under the substrate-potential ansatz
deferred to OP-EHT},
companion paper (2026).

\bibitem{TGP-nbody}
M.~Serafin,
\emph{N-body dynamics in TGP: chaos suppression and Earnshaw violation},
companion paper (2026).

\bibitem{Will2014}
C.~M.~Will,
\emph{The confrontation between general relativity and experiment},
Living Rev.\ Relativity \textbf{17}, 4 (2014).

\bibitem{BertottiCassini}
B.~Bertotti, L.~Iess, and P.~Tortora,
\emph{A test of general relativity using radio links with the Cassini spacecraft},
Nature \textbf{425}, 374 (2003).

\bibitem{WilliamsLLR}
J.~G.~Williams, S.~G.~Turyshev, and D.~H.~Boggs,
\emph{Progress in lunar laser ranging tests of relativistic gravity},
Phys.\ Rev.\ Lett.\ \textbf{93}, 261101 (2004).

\bibitem{Planck2018}
Planck Collaboration,
\emph{Planck 2018 results.\ VI.\ Cosmological parameters},
A\&A \textbf{641}, A6 (2020).

\bibitem{GW170817-speed}
LIGO/Virgo Collaboration,
\emph{Gravitational waves and gamma-rays from a binary neutron star merger:
GW170817 and GRB 170817A}, ApJL \textbf{848}, L13 (2017).

\end{thebibliography}

\end{document}
